---
title: 'Final Project 1: Reproducible Report on NYPD Shooting Incident Data (Historic)'
author: "Karin Rosvall-Clausen"
date: "2024-08-17"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#the below required libraries might need to be installed prior to knitting the file
library("tidyverse")
library("ggthemes")
library("lubridate")
```

### Summary NYPD Shooting Incident Data (Historic)

The NYPD Shooting Incident Data (Historic) dataset includes information about every shooting incident that occurred in NYC from 2006 through the end of the previous calendar year.  The data is manually collected and reviewed by the Office of Management Analysis and Planning prior to being posted on the NYPD website.  As of August 17, 2024, the data was last updated on 4/23/2024.  For more details, you can visit the specific "about_data" webpage at: <https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8/about_data>. 

### Import Data

```{r get_nypd_shooting_dataset, message=FALSE}
nypd_shooting_dataset <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv")
```

### Review Data

1. review the imported data tibble
```{r review_tibble_initial}
nypd_shooting_dataset
```

2. glimpse the imported data
```{r glimpse_data}
glimpse(nypd_shooting_dataset)
```

3. view the imported data
```{r view_data}
view(nypd_shooting_dataset)
```

### What do I want to learn about the NYPD Shooting Incident Data (Historic)?

1. How many total shootings occurred in each Borough and show the split between how many did and did not result in death (statistical_murder_flag = true)?
2. How many total shootings occurred in each Borough by year?

### Tidy and Transform data

1. Create a new dataset with only the below columns, converted the OCCUR_DATE from chr to date, and removed duplicate rows:

  + INCIDENT_KEY
  + OCCUR_DATE - CONVERT FROM CHR > DATE
  + BORO
  + STATISTICAL_MURDER_FLAG

```{r create_new_dataset}
clean_victim_dataset <- nypd_shooting_dataset |>
  mutate(
    DATE = mdy(OCCUR_DATE)) |>
  select(INCIDENT_KEY, DATE, BORO, STATISTICAL_MURDER_FLAG) |>
  distinct()
```

2. review summary of the NEW dataset
```{r review_summary_new_data}
summary(clean_victim_dataset)
```

3. view the NEW dataset
```{r view_summary_new_data}
view(clean_victim_dataset)
```
After a quick review by sorting the data, it appears all the columns have "valid" data.

4. check each remaining column for null data
```{r check_dataset_for_null_values, results='hide'}
which(is.na(clean_victim_dataset$DATE))
which(is.na(clean_victim_dataset$BORO))
which(is.na(clean_victim_dataset$STATISTICAL_MURDER_FLAG))
```
There is no missing (NA) data in the columns

### Visualize and Analyze data

#### Question 1:
How many total shootings occurred in each Borough and show the split between how many did and did not result in death (statistical_murder_flag = true)

```{r plot_default_data, echo=FALSE}
min_date <- format(min(clean_victim_dataset$DATE), format = "%m/%d/%Y")
max_date <- format(max(clean_victim_dataset$DATE), format = "%m/%d/%Y")
subtitle <- str_c("From: ", min_date, " - ", max_date)
```

```{r plot 1, echo=FALSE, warning=FALSE}
clean_victim_dataset |>
  ggplot(aes(y = fct_infreq(BORO), fill = STATISTICAL_MURDER_FLAG)) +
  geom_bar() +
  geom_text(
    aes(label=..count..),
    vjust = -0.5, size = 3,
    stat='count',
    position = position_dodge(width = 1)
  ) +
  labs(
    title = "Total number of shooting incidents in NYC by Borough",
    subtitle = subtitle,
    x = "Total shooting incidents", y = "Borough")
```

##### Analysis:

The borough of Brooklyn has the most shooting incidents and deaths of all the boroughs of NYC recorded from 1/1/2006 - 12/29/2023.  Based upon the rendered graph it also appears Brooklyn might have the highest percentage of murders/shootings than any other borough.

```{r total shooting counts, message=FALSE}
daily_shooting_counts <- clean_victim_dataset |>
  group_by(BORO, DATE) |>
  arrange(BORO,DATE) |> 
  summarize(
    daily_shootings = sum(STATISTICAL_MURDER_FLAG == FALSE),
    daily_murders = sum(STATISTICAL_MURDER_FLAG == TRUE),
    daily_shootings_murders = daily_shootings + daily_murders
    )
```

##### Additional Questions:  

1. Which borough has the highest proportion of deaths/shooting incidents?
2. Which has the lowest?

```{r count_check, echo=FALSE}
total_shooting_counts <- daily_shooting_counts |> 
  group_by(BORO) |> 
  summarise(
    total_shootings = sum(daily_shootings),
    total_murders = sum(daily_murders),
    total_shootings_murders = total_shootings + total_murders,
    total_shootings_pct = (total_shootings / total_shootings_murders) * 100,
    total_murders_pct = (total_murders / total_shootings_murders) * 100 
  )
```

```{r all_boro_murder_pcts, echo=FALSE, results='markup'}
total_shooting_counts |> 
  select(BORO, total_murders_pct, total_shootings, total_murders, total_shootings_murders) |> 
  arrange(desc(total_murders_pct))
```

#### Question 2: 
How many total shootings occurred in each Borough by year

```{r yearly_shootings, echo=FALSE, message=FALSE}
yearly_boro_shootings <- clean_victim_dataset |>
  group_by(BORO, year = year(DATE)) |> 
  summarize(count = n())
```

```{r plot 2, echo=FALSE, message=FALSE}
yearly_boro_shootings |>
  ggplot(aes(x = year, y = count, color = BORO)) +
  geom_line() +
  geom_point() +
  labs(
    title = "NYC shooting trends in NYC by Borough",
    subtitle = subtitle,
    x = "Year", y = "Total count"
  )
```

##### Analysis:

It appears most of the boroughs had a significant decline in shootings from 2011 to 2019.  Then in 2020 the number of incidents for all boroughs increased to 2019 levels or more.


```{r highest count per boro}
boro_shootings_2011_2019_2020 <- yearly_boro_shootings  |> 
  filter(year %in% c("2011", "2019", "2020"))  |> 
  pivot_wider(names_from = year, values_from = count, names_prefix = "year_")  |> 
  rename(year_2011 = year_2011, year_2019 = year_2019, year_2020 = year_2020) |> 
  mutate(
    decr_pct_2011_to_2019 = ((year_2011 - year_2019)/year_2011) * -100,
    incr_pct_2019_to_2020 = ((year_2020 - year_2019)/year_2019) * 100,
    .before = 2
    ) |> 
  arrange(decr_pct_2011_to_2019)

boro_shootings_2011_2019_2020
```

##### Additional Questions: 

1. Why did all the boroughs see a significant decline in shootings from 2011 to 2019?
4. Why did shooting incidents spike in 2020 for all boroughs?  Was it Covid related?

### Bias identification

#### Conclusion:
Brooklyn and the Bronx are the most unsafe boroughs of NYC where Staten Island is the safest based upon total recorded shooting incidents since 2006.  

#### Personal Bias:

In the above statement, I used the terms "unsafe" and "safest" and attempted to "mitigate" my personal bias by including the phrase "based on recorded shooting incidents since 2006." However, there are many other factors that can contribute to being "unsafe," such as car accidents, drug overdoses (both illegal and prescription), domestic violence, and robbery. Additionally, Staten Island has the highest percentage of shootings that have resulted in murder compared to the other boroughs. Given this information, might some people consider Staten Island to be the most unsafe borough?
